---
title: "High-Level Engine Structure"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article describes how the package handles and runs games at a high level.  
For a description of how the player interacts with the game, see `vignette("guide")`.  
For a walkthrough of creating a game, see `vignette("snake")`.

## 1. ROM and RAM
To run a game, we'll use two objects: the ROM and RAM.[^1]

[^1]: Named after the equivalent concepts in retro games!

The ROM (`rom.init()`) contains everything static for the game--- code, images, and information about how to draw the game.

The RAM (`ram.init()`) stores everything dynamic--- player coordinates, game timer, current inputs--- and is constantly updated as the game is run.

This structure is particularly convenient for two reasons. For one, because **all game code** is stored in one object (the ROM), games are easy to handle and load--- we just need to make the ROM object and edit it, or load a premade one like `SuperRrio`.

The other convenience is that, by storing all dynamic info in one object (the RAM), we can easily **copy that object** to create a savestate--- a snapshot of the game at a given moment that can be saved and loaded later. This allows the engine to be very robust, as any issues (like an input not registering) can be rolled back (see `vignette("rollback")`) to an earlier savestate and rerun correctly.

## 2. Making and Running a Game

The basic usage of the engine is structured like this:

1. game dev creates and adds game code, art, etc. to ROM object
2. player uses ROM to initialize RAM object for playing
3. player runs the game using this RAM[^2]
4. player starts an additional R session to send inputs to the game

The two R sessions are referred to as the **display session**, where most of the code happens and the game is drawn, and the **input session**, where the player sends inputs to the game.

[^2]: Steps 2 and 3 are usually combined and run together with `quickload()`.

In code, the four steps look like this:
```{r eval=F}
  #1
Game = rom.init({game stuff})
Game$morestuff = {additional game stuff}

  #2
RAM = ram.init(Game)
	
  #3
RAM = ram.run(Game)

  #4 (in separate session/window)
inputs.listen()
```

The game then runs constantly and responds to any inputs by the player in the input session.

From here, the player can `^C` in the display session to interrupt (pause) the game and inspect the RAM object, continue the game, or even edit the ROM.[^3]

[^3]: Note: since the ROM is loaded into the RAM, changes to the ROM will have to pushed to the RAM with `RAM$ROM = newrom`. It's almost always better to just restart the game by rerunning `ram.init()`.

## 3. The Gameloop
Once the RAM is run, it executes the **gameloop** over and over. This is where all the code dedicated to running the engine, drawing the game, running game logic, etc. is called. 

The gameloop involves four distinct systems:

- **Game Code**: Running the game.

- **Inputs**: Interfacing with the player.

- **Timing**: Executing the gameloop at the right pace.

- **Rendering**: Drawing the game to the console.

Each system has a full article (linked in the sections below) either describing or demonstrating the system.

### 3.1 Game Code
`vignette("snake")`

Game code is run in `ROM$custom()` and is made by the game dev. It reads and writes to RAM to do things like respond to player inputs, move characters around, and trigger events like game end.

### 3.2 Inputs
`vignette("inputs")`

The Input system serves to let the player interact with the game live as it runs. It consists of a **listener** to record player inputs and code to interact with the listener, as well as code to turn recorded inputs into something usable by the Game Code.

The documentation for `inputs.get()` has a good description of the input pipeline, from the player pressing keyboard keys to the input being interpreted by game code.

### 3.3 Timing
`vignette("timing")`

The Timing system ensures that the game runs and draws at the correct speed. This involves speeding up or slowing down when the game is ahead or behind of where it "should" be (according to the desired speed, determined by `ROM$framerate`), and deciding whether or not to draw a given tick--- for the game to look smooth, drawing must only happen when the game is "caught up".

### 3.4 Rendering
`vignette("render")`

The Rendering system consists of a host of functions dedicated to rendering the current **scene**---the way objects are organized onscreen on a given tick/gamestate---to the console, which is where the game is displayed.

This involves:

- getting the right sprite (image) for each object and putting it in the right place and layer in the scene (`render.sprite()`)

- combining the whole scene into one object to be rendered (`render.scene()`)

- making this object printable to the console (`render.matrix()`)

## 4. Alternative Methods of Interaction
While the default way of interacting with the game is using `ram.run()` and `inputs.listen()`, a couple other options are available.

### 4.1 Listener Frame Advance
The live game can be suspended at any time by inputting `/pause` in the input session. Then:

* The player can input `/tick` to advance one tick forward (and draw the scene again).

* The player can type regular inputs, which will be processed by the game the next time `/tick` is called (or the game is resumed).

* The player can resume live gameplay by inputting `/resume`.

This behavior is documented in `?inputs.command`. I find this method to be more convenient than that in the next section, since staying in the input session allows for more fluid transition between live and tick-by-tick gameplay.

### 4.2 Manual Advance
Forgoing the need for the listener session, the game can also be paused and manipulated directly in the display session. This gives the user more control over altering the RAM that may be especially useful for debugging.

This is done by pausing the game with `^C` (which ends the `ram.run` process). The RAM object can then be manipulated freely:

* `RAM = ram.tick(RAM)` advances the game one tick.

* `RAM = ram.input(RAM, [input])` adds an input to RAM that will be processed on the next tick.

* `RAM = ram.run(RAM)` resumes the game (requiring the input session again to interact)

While paused, the RAM can also be easily saved and reloaded by copying it, as in the following code:

```{r eval=F}
savestate_1 = RAM
RAM = ram.run(RAM)
#play the game a bit via the input session
^C
RAM = savestate_1 #restore the savestate
RAM = ram.run(RAM) #the game is now back to where it was when savestate_1 was saved
```

## 5. Gameloop Function Tree
The gameloop consists of the following nesting of functions in order:

* `ram.run()` once to start or resume the game

    - **Timing**:

    * `ram.update()` every tick (see `vignette("timing")`)
        
         - **Inputs**:
         
         * `inputs.get()`

            * `inputs.read()` 
        
            * `inputs.command()` if any commands are sent
            
            * `ram.rollback`()` if any inputs were received late

        * `inputs.process()`
        
        - **Game Code**:  

        * `ram.tick()`   

            * `RAM$ROM$custom()`  
            
        - **Rendering**:

        * `render.ram()` if `RAM$time >` `time.sec()`  

            * `render.object()` for every object in `RAM$objects`  
            
                * `render.animate()` if the object's sprite is animated  
            
                * `render.sprite()` or custom `obj$draw()`  
                
                    * `render.overlay()`  

            * `render.scene()`    

                * `render.overlay()` for every layer in `scene$layers`    

            * `render.matrix()`    

This tree can also be found in the documentation in `?ram.run`.
