---
title: "R^2^Studio"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{r2studio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rcade)
```

We have a graphics system. Why not render an R console?

This article describes the `R2Studio` ROM, which lets you run RStudio... in RStudio. Unlike `vignette("snake")`, the article does not use a walkthrough-based format, and instead just covers the various systems present in making the ROM work. You can always inspect the ROM object (`View(R2Studio)`) to take a closer look at its code.


# GO THROUGH SNAKE and organize the source code into nice chunks #----

# 1. Goals

render the regular console (inputs, outputs, etc.)
hook the plot command and do basic scatter/line plots

We don't need a smooth framerate for this--- one frame per second should be plenty.

# 2. UI
The UI is inspired by RStudio and features a console and plotting window. There are a couple different sprites dedicated to drawing the borders, and they generate their sprite live in `obj$draw()`; this lets the UI adapt to changes in display resolution 

[DISPLAY RES CHANGES SUCH AS THOSE CAUSED BY THE SPECIAL FUNCTIONS]

# 3. Console

# 4. Plotting
goals

# 4.1 Hooking `plot()`


# 5. Meta Functions
In addition to hooking `plot()`, I've provided some convenience functions that interact with the game itself to, change the font, resize the display area, etc. All they do is edit `RAM$ROM` properties, and the game is rendered in such a way that respects live changes to these--- almost everything draws through `obj$draw()` rather than the default sprite system.

note: hooks won't work for the syntax `plot ()`

# what are these meta functions? list them all and have a small image demo (e.g. to compare fonts and sizes)



# R^3^Studio




